---
title: "Analysing cultural frequency data: neutral theory
and beyond"
author: "Enrico Crema"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: true
    self_contained: yes
fontsize: 11pt
documentclass: article
vignette: >
  %\VignetteIndexEntry{Analysing cultural frequency data: neutral theory and beyond}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction and Setup

This markdown document contains the R script for reproducing some of the figures of the book chapter "Analysing cultural frequency data: neutral theory and beyond" by A.Kandler and E.Crema. The main functions for running the simulations described in the chapter can be installed as an R package with following command:

```{r,installpackage,eval=FALSE}
devtools::install_github("ercrema/HERAChp.KandlerCrema")
```

In order to fully reproduce the figures in the text the following R packages must also be loaded:

```{r,load_packages,message=FALSE,results='hide'}
library(HERAChp.KandlerCrema)
library(plotrix)
library(LaplacesDemon)
library(ggridges)
library(ggplot2)
library(foreach)
library(doParallel)
```

Some of the figures require a considerable number of simulation runs. In order to reduce running time the example belows are based on 100 repetitions instead of the 1,000 used in the book chapter, and most simulations are executed using multiple threads. 

```{r set_nsim}
nsim = 100 #number of repetitions
ncores = detectCores()-1 #use total number of available threads minus 1
```

Notice that even with the default settings above executing all the scripts below took about 40 minutes using a machine with with 8 logical CPUs (3.3 GHz Intel Core i7) and 16 GB Ram machine.  

# Figure 1: Diversity and Battleship Curves

The function `transmission()` executes unbiased and frequency biased transmission returning a variety of statistics. The example below compares the time-series of the Simpson's diversity index against the the so-called battle-ship curve of the most common variants. First we run the model and extract relevant outputs:


```{r,fig1.sim,fig.height=5,fig.width=8}
#set random seed
set.seed(123)
#run simulation
res.fig1=transmission(timesteps=2000,mu=0.005,N=500,warmUp=1000,bias=0,raw=TRUE)
#extract matrix of frequencies
mat.fig1=res.fig1$rawMatrix
#compute proportions
pmat.fig1=prop.table(mat.fig1,1)
#reorder matrix based on total frequencies
pmat.fig1=pmat.fig1[,order(apply(pmat.fig1,2,sum),decreasing=TRUE)]
#extract observed diversity time-series
divTS = res.fig1$obs.div
#extract expected diversity under neutrality
divExp = res.fig1$exp.div
```

The matrix `pmat.fig1` now contains the frequencies of each variant at each timestep, the vector `divTS` contains the time-series of Simpson's diversity index, while `divExp` contains the expected diversity under neutrality, equivalent to $1-\frac{1}{2N\mu+1}$. We are now ready to make our figure:


```{r,fig1.plot,fig.height=5,fig.width=8}
#make figure
par(mfrow=c(1,2))

#left panel 
par(mar=c(5,0,5,0))
topVariants = 10 #show only the 10 most frequent variants
battleship.plot(pmat.fig1[,1:topVariants],yaxlab="",xaxlab= letters[1:topVariants],border=NA,col="grey",mar=c(5,5,5,1))
axis(2,padj=0.5)
mtext(side=2,"time",line=2)
mtext(side=3,paste0("Top ",topVariants," variants"),line=1)
#right panel
par(mar=c(5,0,5,0))
plot(divTS,1:1000,type="l",xlab="",axes=F,ylab="",xlim=c(0,1))
abline(v=divExp,lty=2)
axis(side=1,padj=-0.5)
mtext(side=1,expression(1-f[t]),line=2)
mtext(side=3,"Diversity ",line=1)
```

# Figure 2: Effect of sampling window for computing reliable estimates of the turn-over rate

Here we explore how the turn-over rate based estimate ofthe exponent $x$ is affected by the size (duration) of the sampling window. We compare the expected value f $x$ under neutrality (0.86) against distributions of simulation outputs with different number of timesteps (i.e. differently sized of sampling window). We first execute the simulation (notice that with `nsim=1000` the process will take XXX hours) 


```{r,fig2.simulation}

#set sampling window from 10 to 500, all with 1000 generations of warmup
timesteps <- c(seq(1010,1100,10),1200,1500)

# set parameter space
timesteps.param = rep(timesteps,each=nsim)

#set up parallel computing
cl <-  makeCluster(ncores)  
registerDoParallel(cl)

# execute simulation in parallel and extract estimates of the exponent x (here named b):
res.fig2.vector = foreach(i=1:length(timesteps.param),.combine=c,.export="transmission") %dopar% {
  tmp=transmission(timesteps=timesteps.param[i],mu=0.01,N=500,warmUp=1000,bias=0,raw=F,top=10)$b
}

#stop cluster
stopCluster(cl)

# convert output into a matrix
res.fig2 = matrix(res.fig2.vector,nrow=nsim,ncol=length(timesteps))

#Extract summary statistics (median, first and third quartile) from each setting of the sampling window
mid.fig2=apply(res.fig2,2,median)
lo.fig2=apply(res.fig2,2,quantile,0.25)
up.fig2=apply(res.fig2,2,quantile,0.75)
```

We can now visualise our output. The object `res.fig2` contains the estimates of $x$ under different sampling window size, so a quick way to plot the simulation output is to just make a box and whisker plot:

```{r,fig2.plot1,fig.width=9,fig.height=6}
boxplot(res.fig2,names=timesteps-1000,xlab="time-steps",ylab="Estimates of x",col="bisque3")
abline(h=0.86,col=2,lty=2) #expected value of x
```


The code below is instead the one used in the manuscript:

```{r,fig2.plot2,fig.width=9,fig.height=6}
#Plot results using rect() and unqeual spacing along the x-axis
par(mar=c(5,4,4,1))
plot(1:14,1:14,ylim=c(min(lo.fig2),max(up.fig2)),type="n",axes=F,xlab="N. of Timesteps",ylab="estimates of x")

for (x in 1:10)
{
	rect(xleft=x-0.25,xright=x+0.25,ybottom=lo.fig2[x],ytop=up.fig2[x],col="lightgrey",border=NA)
	lines(x=c(x-0.25,x+0.25),y=c(mid.fig2[x],mid.fig2[x]),lwd=2)
}

x=11
rect(xleft=12-0.25,xright=12+0.25,ybottom=lo.fig2[x],ytop=up.fig2[x],col="lightgrey",border=NA)
lines(x=c(12-0.25,12+0.25),y=c(mid.fig2[x],mid.fig2[x]),lwd=2)

x=12
rect(xleft=14-0.25,xright=14+0.25,ybottom=lo.fig2[x],ytop=up.fig2[x],col="lightgrey",border=NA)
lines(x=c(14-0.25,14+0.25),y=c(mid.fig2[x],mid.fig2[x]),lwd=2)

# add axes
axis(side=2,at=seq(0,5,0.5),cex.axis=0.8,las=2)
axis(side=1, at=c(1:10,12,14),labels=c(seq(10,100,10),200,500),cex=0.9)

# show the expected value of b under neutrality
abline(h=0.86,lty=2,col="red")

```

# Figure 3: Effect of a Heterogenous population of learners

To execute a simulation with an heterogenous population of learners we use the function `heteroPopTransmission()` which enables to define the frequency bias of each individual at each generation as a random draw from a normal distribution with a user defined mean (argument `bmean`) and standard deviation (argument `bsd`). Here we explore a scenario where the mean is hold constant to 0 (hence on average the learners engage into an unbiased transmission), but the standard deviation is varied and explored. 


```{r,fig3.simulation}
#set number of generations
ngen=1000
#set parameter to vary (bsd)
bsd = c(0,0.1,0.2)
#set parameter space
bsd.param = rep(bsd,each=nsim)

#set up parallel computing
cl <-  makeCluster(ncores)  
registerDoParallel(cl)

res.fig3.vector = foreach(i=1:length(bsd.param),.combine=c,.export="heteroPopTransmission") %dopar% {
  heteroPopTransmission(N=500,bmean=0,bsd=bsd.param[i],mu=0.01,timesteps=ngen)[ngen]
}

#stop cluster
stopCluster(cl)

# convert output into a matrix
res.fig3 = matrix(res.fig3.vector,nrow=nsim,ncol=length(bsd))
```


Each column of the matrix `res.fig3` now contains the result of 1,000 simulation runs with different levels of hetogeneity in the population of learners as defined by the parameter `bsd`.

```{r,fig3.plot,fig.height=4,fig.width=12}
# Setup layout
mat=matrix(c(1,4,4,4,2,5,5,5,3,6,6,6),nrow=2,ncol=6)
layout(mat=mat,heights=c(0.45,0.55),widths=c(0.45,0.55,0.45,0.55,0.45,0.55))

# Visualise distribution of learners' frequency bias under each setting
par(mar=c(5,5,1,1))
plot(0,1,type="n",axes=F,xlab="",ylab="",xlim=c(-0.6,0.6))
lines(x=c(0,0),y=c(0,2),lwd=2)
axis(1,at=c(-0.4,0,0.4),padj=-0.5)
mtext("b",1,line=2,cex=0.8)

par(mar=c(5,5,1,1))
xx=seq(-0.6,0.6,length.out=1000)
yy=dnorm(x=xx,sd=0.1)
plot(xx,yy,type="n",axes=F,xlab="",ylab="")
polygon(c(xx,rev(xx)),c(yy,rep(0,1000)),col="black")
axis(1,at=c(-0.4,0,0.4),padj=-0.5)
mtext("b",1,line=2,cex=0.8)

par(mar=c(5,5,1,1))
xx=seq(-0.6,0.6,length.out=1000)
yy=dnorm(x=xx,sd=0.2)
plot(xx,yy,type="n",axes=F,xlab="",ylab="")
polygon(c(xx,rev(xx)),c(yy,rep(0,1000)),col="black")
axis(1,at=c(-0.4,0,0.4),padj=-0.5)
mtext("b",1,line=2,cex=0.8)


# Show simulation result
par(mar=c(5,5,1,1.2))
hist(res.fig3[,1],breaks=seq(0.55,1,0.02),border=NA,col="bisque2",xlab="Diversity",main="",cex.lab=1.5,cex.axis=1.3)
abline(v=1-(1/(500* 0.01*2+1)),lty=2,col=2,lwd=1.5)
abline(v=mean(res.fig3[,1]),lty=3,lwd=1.5)
legend("left",legend=c("Average Diversity (observed)","Expected Diversity"),col=c(1,2),lty=c(3,2),bty="n",cex=0.8,y.intersp=1.5)

hist(res.fig3[,2],breaks=seq(0.55,1,0.02),border=NA,col="bisque2",xlab="Diversity",main="",cex.lab=1.5,cex.axis=1.3)
abline(v=1-(1/(500* 0.01*2+1)),lty=2,col=2,lwd=1.5)
abline(v=mean(res.fig3[,2]),lty=3,lwd=1.5)

hist(res.fig3[,3],breaks=seq(0.55,1,0.02),border=NA,col="bisque2",xlab="Diversity",main="",cex.lab=1.5,cex.axis=1.3)
abline(v=1-(1/(500* 0.01*2+1)),lty=2,col=2,lwd=1.5)
abline(v=mean(res.fig3[,3]),lty=3,lwd=1.5)
```


# Figure 4: Effect of non-equilibrium

The argument `bias` in the function `transmission()` allows for a time-varying frequency bias. The code below explores how a temporary shift from an unbiased transmission to an anti-conformist transmission affects the observed diversity of the cultural variants. 

```{r, open_clusters_fig4, include=FALSE}

```


```{r}
#define bias parameter
bias.eq = 0
bias.neq = c(rep(0,1500),rep(0.5,10),rep(0,490)) 


#run simulations
#res.fig4.eq=matrix(replicate(transmission(timesteps=2000,mu=0.005,N=500,warmUp=1000,bias=bias.eq,raw=TRUE)$obs.div[401:900],n=nsim),nrow=500,ncol=nsim)

#set up parallel computing
cl <-  makeCluster(detectCores()-1)  #use one core less than the total available
registerDoParallel(cl)

res.fig4.eq.vector = foreach(i=1:nsim,.combine=c,.export="transmission") %dopar% {
  transmission(timesteps=2000,mu=0.005,N=500,warmUp=1000,bias=bias.eq,raw=TRUE)$obs.div[401:900]
}

res.fig4.neq.vector = foreach(i=1:nsim,.combine=c,.export="transmission") %dopar% {
  transmission(timesteps=2000,mu=0.005,N=500,warmUp=1000,bias=bias.neq,raw=TRUE)$obs.div[401:900]
}

#stop cluster
stopCluster(cl)

res.fig4.eq = matrix(res.fig4.eq.vector,nrow=500,ncol=nsim)
res.fig4.neq = matrix(res.fig4.neq.vector,nrow=500,ncol=nsim)

# Compute Expected Diversity
expected.diversity = 1 - (1/(2*500*0.005+1))
```


```{r,fig4.plot1,fig.width=8,fig.height=6}
par(mfrow=c(2,1),mar=c(1,5,5,2))

plot(1:500,res.fig4.eq[,1],ylim=c(0.4,1),type="n",xlab="",ylab="Simpson's Diversity",xaxs="i",axes=F,las=2)
text(480,0.5,"a",cex=1.5)
polygon(x=c(1:500,500:1),y=c(apply(res.fig4.eq,1,quantile,prob=0.025),rev(apply(res.fig4.eq,1,quantile,prob=0.975))),col=rgb(1,0,0,0.2),border=NA)
apply(res.fig4.eq,2,lines,col=rgb(0,0,0,0.01),x=1:500)
lines(1:500,apply(res.fig4.eq,1,mean),col="red")
abline(h=expected.diversity,col="black",lty=2)
axis(2,at=seq(0,1,0.2),las=2,hadj=0.8)

par(mar=c(6,5,0,2))

plot(1:500,res.fig4.neq[,1],ylim=c(0.4,1),type="n",xlab="time",ylab="Simpson's Diversity",xaxs="i",axes=F,las=2)
polygon(x=c(1:500,500:1),y=c(apply(res.fig4.neq,1,quantile,prob=0.025),rev(apply(res.fig4.neq,1,quantile,prob=0.975))),col=rgb(1,0,0,0.2),border=NA)
apply(res.fig4.neq,2,lines,col=rgb(0,0,0,0.01),x=1:500)
lines(1:500,apply(res.fig4.neq,1,mean),col="red")
abline(h=expected.diversity,col="black",lty=2)
rect(xleft=100,xright=110,ybottom=-100,ytop=100,col=rgb(0,0,1,0.2),lty=3)
text(480,0.5,"b",cex=1.5)
axis(2,at=seq(0,1,0.2),las=2,hadj=0.8)
axis(1,at=c(1,100,200,300,400,500))
```


# Figure 7: Posterior Distribution of the Merzbach Assemblage Analysis.

The code below plots the results of the 

```{r,fig7.load}
# load data
data("post_b_merzbach")
```

```{r,fig7.plot1,fig.width=6,fig.height=8}
# Simple boxplot:
par(mar=c(5,8,1,1))
boxplot(post_b_merzbach,horizontal=T,las=2,xlab="b")
abline(v=0,lty=2,col=2) #unbiased transmission
```


```{r,fig7.plot2,fig.width=6,fig.height=8}
#Extract 50 and 95$ HPDI
hpd50 = t(apply(post_b_merzbach,2,p.interval,prob=0.5))
hpd95 = t(apply(post_b_merzbach,2,p.interval,prob=0.95))


#Restructure into data.frame
data=rbind.data.frame(
data.frame(Model="Equilibrium",b=post_b_merzbach[,1],Phase="All(Eq)"),
data.frame(Model="Var.Population",b=post_b_merzbach[,2],Phase="All(S.Eq)"),
data.frame(Model="Var.Pop & Transmission",Phase="VIII",b=post_b_merzbach[,3]),
data.frame(Model="Var.Pop & Transmission",Phase="IX",b=post_b_merzbach[,4]),
data.frame(Model="Var.Pop & Transmission",Phase="X",b=post_b_merzbach[,5]),
data.frame(Model="Var.Pop & Transmission",Phase="XI",b=post_b_merzbach[,6]),
data.frame(Model="Var.Pop & Transmission",Phase="XII",b=post_b_merzbach[,7]),
data.frame(Model="Var.Pop & Transmission",Phase="XIII",b=post_b_merzbach[,8]),
data.frame(Model="Var.Pop & Transmission",Phase="XIV",b=post_b_merzbach[,9]))

#make joyplot
g=ggplot(data, aes(x = b, y = Phase,fill=Model)) + 
geom_density_ridges(rel_min_height=0.005,alpha=0.5,color="white")+scale_fill_manual(values=c("indianred","royalblue","darkgrey"))+
  theme_ridges(center_axis_labels = TRUE)+
    theme(plot.margin = unit(c(1,1,1,1), "cm"),legend.position = c(0.6, 0.8))+scale_y_discrete(labels=c("All","All","VIII","IX","X","XI","XII","XIII","XIV"))+coord_cartesian(xlim = c(-0.25, 0.25)) +  scale_x_continuous(expand = c(0.01, 0))+
    annotate("rect",xmin = 0.05, xmax = 2.5, ymin = 6.5, ymax = 8.7,alpha=0.7,fill="white")+
           annotate("segment",x = 0.06, xend = 0.077, y = 7.1, yend = 7.1)+
           annotate("segment",x = 0.06, xend = 0.06, y = 7.1, yend = 7.05)+
           annotate("segment",x = 0.077, xend = 0.077, y = 7.1, yend = 7.05)+
            annotate("segment",x = 0.06, xend = 0.06, y = 7.1, yend = 7.15)+
           annotate("segment",x = 0.077, xend = 0.077, y = 7.1, yend = 7.15)+
           annotate("text",x = 0.12,y = 7.1,label="50% HPDI")+
           annotate("segment",x = 0.06, xend = 0.078, y = 6.8, yend = 6.8,linetype="dotted")+
           annotate("text",x = 0.12,y = 6.8,label="95% HPDI")
        
#add HPDI intervals  
  for (i in 1:9)
  {
    g = g +
    annotate("segment",x = hpd50[i,1], xend = hpd50[i,2], y = i+0.3, yend = i+0.3,lwd=0.5)+
    annotate("segment",x = hpd50[i,1], xend = hpd50[i,1], y = i+0.3, yend = i+0.25,lwd=0.5)+
    annotate("segment",x = hpd50[i,2], xend = hpd50[i,2], y = i+0.3, yend = i+0.25,lwd=0.5)+ 
    annotate("segment",x = hpd50[i,1], xend = hpd50[i,1], y = i+0.3, yend = i+0.35,lwd=0.5)+
    annotate("segment",x = hpd50[i,2], xend = hpd50[i,2], y = i+0.3, yend = i+0.35,lwd=0.5)+
    annotate("segment",x = hpd95[i,1], xend = hpd95[i,2], y = i+0.3, yend = i+0.3, linetype="dotted",lwd=0.5)
  }
g
```